---
title: "Clinical Tables with {gtsummary}"
subtitle: "Hands-on Workshop — Participant Workbook"
author: "Clinical Data Programming"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-tools: true
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  eval: false
editor: visual
---

```{r}
#| label: setup
#| include: false
#| eval: true

knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

# How This Workbook Works {#sec-how}

Every exercise gives you three things and withholds a fourth.

| You get                                              | You don't get      |
|------------------------------------|------------------------------------|
| A **task** stating what the table must show          | Working code       |
| A **skeleton** with blanks marked `____`             | The completed call |
| **Check your work** — facts your output must satisfy | A solution chunk   |

Hints are collapsed. Open the first one only after you have tried something; open the second only if the first did not unstick you. There is no third hint on purpose.

::: callout-important
## Fill in the blanks, then remove `#| eval: false`

Code chunks in this workbook are set not to run, so the document renders even with blanks in it. When you have completed a chunk, delete its `#| eval: false` line and run it with `Ctrl/Cmd + Enter`.
:::

## Environment

Run this first. If anything fails, sort it out before moving on.

```{r}
#| label: packages
#| eval: true

library(gtsummary)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(gt)
library(pharmaverseadam)

packageVersion("gtsummary")   # must be 2.0.0 or later
```

::: callout-warning
## If `pharmaverseadam` will not install

Substitute your own ADSL / ADAE / ADLB. Every exercise uses standard CDISC variable names, so the only change needed is the treatment variable if yours differs.
:::

```{r}
#| label: theme
#| eval: true

theme_gtsummary_compact()
```

## ------------------------------------------------------------------------

# Part 1 — Demographics and Baseline Characteristics {#sec-demog}

## Prepare the data

Run this as given.

```{r}
#| label: demog-data
#| eval: true

adsl <- pharmaverseadam::adsl |>
  filter(SAFFL == "Y") |>
  mutate(
    TRT01A = factor(
      TRT01A,
      levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
    ),
    SEX = factor(SEX, levels = c("M", "F"), labels = c("Male", "Female")),
    AGEGR1 = factor(AGEGR1, levels = c("<65", "65-80", ">80"))
  )

count(adsl, TRT01A)
```

## Exercise 1.1 — The shell

**Task.** Treatment across the columns. Rows: `AGE`, `AGEGR1`, `SEX`, `RACE`, `ETHNIC`.

```{r}
#| label: ex-1-1
#| eval: false

adsl |>
  tbl_summary(
    by = ____,
    include = c(____)
  )
```

## Exercise 1.2 — Statistics the SAP asks for

**Task.** Report `AGE` on four separate lines — n, Mean (SD), Median, Min – Max — and categorical variables as n (%).

```{r}
#| label: ex-1-2
#| eval: false

adsl |>
  tbl_summary(
    by = TRT01A,
    include = c(AGE, AGEGR1, SEX, RACE, ETHNIC),
    type = list(AGE ~ "____"),
    statistic = list(
      AGE ~ c("____", "____", "____", "____"),
      all_categorical() ~ "____"
    )
  )
```

## Exercise 1.3 — Decimal places

**Task.** n as integer, mean and SD to one decimal, median to one decimal, min and max as integers. Categorical percentages to one decimal.

```{r}
#| label: ex-1-3
#| eval: false

adsl |>
  tbl_summary(
    by = TRT01A,
    include = c(AGE, AGEGR1, SEX, RACE, ETHNIC),
    type = list(AGE ~ "continuous2"),
    statistic = list(
      AGE ~ c("{N_nonmiss}", "{mean} ({sd})", "{median}", "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      AGE ~ list(____, ____, ____, ____),
      all_categorical() ~ c(____, ____)
    )
  )
```

## Exercise 1.4 — Presentation

**Task.**

1.  Row labels: "Age (years)", "Age Group", "Sex", "Race", "Ethnicity"
2.  First column header: "Characteristic"
3.  Treatment headers showing the arm name, then `(N = xx)` on a second line
4.  Variable labels in bold
5.  Suppress the missing rows entirely

Save the result as `tbl_demog` — later exercises use it.

```{r}
#| label: ex-1-4
#| eval: false

tbl_demog <- adsl |>
  tbl_summary(
    by = TRT01A,
    include = c(AGE, AGEGR1, SEX, RACE, ETHNIC),
    type = list(AGE ~ "continuous2"),
    statistic = list(
      AGE ~ c("{N_nonmiss}", "{mean} ({sd})", "{median}", "{min} - {max}"),
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      AGE ~ list(0, c(1, 1), 1, c(0, 0)),
      all_categorical() ~ c(0, 1)
    ),
    label = list(
      ____ ~ "____",
      # ... one line per variable
    ),
    missing = "____"
  ) |>
  modify_header(
    label = "**____**",
    all_stat_cols() ~ "**____**  \n(N = ____)"
  ) |>
  ____()

tbl_demog
```

## Exercise 1.5 — Total column and footnotes

**Task.** Add a Total column on the **right-hand** side, give the table a caption naming it Table 14.1.2, remove the automatic footnote, and add your own explaining the percentage denominator.

```{r}
#| label: ex-1-5
#| eval: false

tbl_demog_final <- tbl_demog |>
  add_overall(
    last = ____,
    col_label = "____"
  ) |>
  modify_caption("____") |>
  ____(columns = everything()) |>
  modify_source_note("____")

tbl_demog_final
```

------------------------------------------------------------------------

# Part 2 — Adverse Events by SOC and Preferred Term {#sec-ae}

## Prepare the data

Run this as given.

```{r}
#| label: ae-data
#| eval: true

adae <- pharmaverseadam::adae |>
  filter(SAFFL == "Y", TRTEMFL == "Y") |>
  mutate(
    TRTA = factor(
      TRT01A,
      levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
    )
  )

adsl_safety <- adsl |>
  select(USUBJID, TRT01A) |>
  rename(TRTA = TRT01A)

adae |>
  summarise(
    n_records  = n(),
    n_subjects = n_distinct(USUBJID),
    n_soc      = n_distinct(AEBODSYS),
    n_pt       = n_distinct(AEDECOD)
  )
```

## Exercise 2.1 — Counting subjects

**Task.** A flat table of preferred terms — no SOC nesting yet — showing n (%) of subjects by arm, restricted to terms occurring in more than 10 subjects overall, sorted most frequent first. Include a top row counting subjects with any TEAE.

```{r}
#| label: ex-2-1
#| eval: false

adae |>
  tbl_hierarchical(
    variables = ____,
    by = ____,
    id = ____,
    denominator = ____,
    overall_row = ____
  ) |>
  sort_hierarchical(sort = "____") 
```

## Exercise 2.2 — Nest PT inside SOC

**Task.** Same table, now with preferred terms nested under system organ class. SOC rows count subjects with any event in that SOC; PT rows count subjects with that term. Statistics as n (%) with percentages to one decimal.

Save as `tbl_ae`.

```{r}
#| label: ex-2-2
#| eval: false

tbl_ae <- adae |>
  tbl_hierarchical(
    variables = c(____, ____),
    by = TRTA,
    id = USUBJID,
    denominator = adsl_safety,
    overall_row = TRUE,
    statistic = everything() ~ "____",
    digits = everything() ~ list(n = ____, p = ____)
  )

tbl_ae
```

## Exercise 2.3 — Sort and filter

**Task.** Apply the two rules nearly every AE shell requires:

1.  SOCs sorted by descending overall frequency, PTs sorted by descending frequency within their SOC
2.  Only preferred terms occurring in at least 5% of subjects in **any** treatment arm

```{r}
#| label: ex-2-3
#| eval: false

tbl_ae_sorted <- tbl_ae |>
  sort_hierarchical(____) |>
  filter_hierarchical(____)

tbl_ae_sorted
```

## Exercise 2.4 — Presentation

**Task.** Headers with arm and N. First column header showing SOC and PT on two lines. Caption as Table 14.3.1. SOC rows bold. A footnote covering the counting rule, the sort order, the 5% cut-off, and the MedDRA version.

```{r}
#| label: ex-2-4
#| eval: false

tbl_ae_final <- tbl_ae_sorted |>
  modify_header(
    label = "____",
    all_stat_cols() ~ "____"
  ) |>
  modify_caption("____") |>
  ____() |>
  modify_source_note("____")

tbl_ae_final
```

## Exercise 2.5 — Extension

Pick one. These are deliberately less scaffolded.

**(a)** The same table restricted to serious adverse events.

**(b)** An AE table by **maximum severity per subject** — each subject counted once per preferred term, at their worst grade across all episodes of that term.

**(c)** A table showing subject counts and event counts side by side.

```{r}
#| label: ex-2-5
#| eval: false

# Your code here
```

------------------------------------------------------------------------

# Part 3 — Laboratory Results by Visit {#sec-lab}

The target is the shell circulated before the session: visits as row groups, treatment arms spanning the columns, and **Value at Visit** beside **Change from Baseline** within each arm.

## The strategy

There is more than one way to reach this layout. The one you will build here goes:

1.  Summarise `AVAL` for a single visit → one table
2.  Summarise `CHG` for the same visit → a second table
3.  `tbl_merge()` the pair → six statistic columns, in the wrong order
4.  Repair the column order, the headers, and the spanning headers
5.  Blank the change cells at Baseline
6.  Wrap all of that in a function, `map()` it over the visit list, and `tbl_stack()` the results

::: callout-note
## Why this route

Steps 1 to 5 operate on **one visit at a time**. That means you can look at a single small table, see exactly what is wrong with it, and fix it — before the function ever runs over the full visit list. Approaches that build all visits in one call are shorter to write and considerably harder to debug when a column lands in the wrong place.

It also means the function you end up with is testable. That matters more than elegance.
:::

## Prepare the data

Run this as given.

```{r}
#| label: lab-data
#| eval: true

adlb <- pharmaverseadam::adlb |>
  filter(SAFFL == "Y", PARAMCD == "ALT") |>
  mutate(
    TRTA = factor(
      TRTA,
      levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
    )
  )

adlb |>
  distinct(AVISITN, AVISIT) |>
  arrange(AVISITN)
```

## Exercise 3.1 — One visit, value only

**Task.** Summarise `AVAL` at Week 12 by treatment arm: n, Mean (SD), Median, Min – Max, all to two decimals. Label the variable row with the visit name rather than "AVAL".

```{r}
#| label: ex-3-1
#| eval: false

visit <- "Week 12"

df_visit <- adlb |> filter(AVISIT == ____)

tbl_aval <- df_visit |>
  tbl_summary(
    by = ____,
    include = ____,
    type = AVAL ~ "____",
    statistic = AVAL ~ c("____", "____", "____", "____"),
    digits = AVAL ~ list(mean = ____, sd = ____, median = ____, min = ____, max = ____),
    label = list(AVAL ~ ____),
    missing = "____"
  )

tbl_aval
```

## Exercise 3.2 — The same visit, change from baseline

**Task.** Repeat for `CHG`, with the same statistics, the same digits, and the same label.

```{r}
#| label: ex-3-2
#| eval: false

tbl_chg <- df_visit |>
  tbl_summary(
    # your call from 3.1, for CHG
  )

tbl_chg
```

## Exercise 3.3 — Merge the pair

**Task.** Join the two tables side by side, with "Value at Visit" and "Change from Baseline" as spanning headers.

```{r}
#| label: ex-3-3
#| eval: false

tbl2 <- tbl_merge(
  list(____, ____),
  merge_vars = c("var_type", "var_label", "row_type", "label"),
  tab_spanner = c("____", "____")
)

tbl2

# Look at what you got
tbl2$table_body |> names()
```

## Exercise 3.4 — Fix the column order

**Task.** The shell wants value and change **adjacent within each arm**. Reorder the columns of `table_body` to interleave them.

```{r}
#| label: ex-3-4
#| eval: false

new_order <- c("label", "____", "____", "____", "____", "____", "____")

tbl2$table_body <- tbl2$table_body |>
  dplyr::relocate(dplyr::all_of(new_order))

tbl2
```

## Exercise 3.5 — Headers and spanning headers

**Task.** The shell puts the **treatment arm** on top and **value / change** beneath it. You currently have the reverse. Swap them.

```{r}
#| label: ex-3-5
#| eval: false

tbl3 <- tbl2 |>
  modify_header(
    stat_1_1 ~ "____", stat_1_2 ~ "____",
    stat_2_1 ~ "____", stat_2_2 ~ "____",
    stat_3_1 ~ "____", stat_3_2 ~ "____"
  ) |>
  modify_spanning_header(
    c(stat_1_1, stat_1_2) ~ ____,
    c(stat_2_1, stat_2_2) ~ ____,
    c(stat_3_1, stat_3_2) ~ ____
  )

tbl3
```

## Exercise 3.6 — Blank the baseline change cells

**Task.** Change from baseline is undefined at Baseline. Replace those cells with a blank.

```{r}
#| label: ex-3-6
#| eval: false

tbl3$table_body <- tbl3$table_body |>
  mutate(across(c(____, ____, ____), ~ ifelse(____ == "Baseline", " ", .x)))

tbl3
```

## Exercise 3.7 — Wrap it in a function

**Task.** Everything from 3.1 to 3.6 handles one visit. Turn it into a function taking a single argument, then apply it across visits and stack the results.

```{r}
#| label: ex-3-7-fun
#| eval: false

build_visit_tbl <- function(____) {

  df_visit <- adlb |> filter(AVISIT == ____)

  tbl_aval <- df_visit |> tbl_summary( ____ )

  tbl_chg  <- df_visit |> tbl_summary( ____ )

  tbl2 <- tbl_merge( ____ )

  new_order <- c( ____ )
  tbl2$table_body <- tbl2$table_body |> dplyr::relocate(dplyr::all_of(new_order))

  tbl3 <- tbl2 |>
    modify_header( ____ ) |>
    modify_spanning_header( ____ )

  tbl3$table_body <- tbl3$table_body |> mutate( ____ )

  return(____)
}
```

**Task, continued.** Build the visit list. Exclude unscheduled visits, early-removal visits, and any generic post-baseline label.

```{r}
#| label: ex-3-7-visits
#| eval: false

visits <- adlb |>
  filter(
    !str_detect(AVISIT, "____"),
    !str_detect(AVISIT, "____"),
    !str_detect(AVISIT, "____")
  ) |>
  distinct(AVISIT) |>
  pull()

visits
```

**Task, continued.** Run it over the first four visits and stack.

```{r}
#| label: ex-3-7-map
#| eval: false

visit_tbls <- map(visits[1:4], ____)

tbl_final <- tbl_stack(____)

tbl_final
```

# Part 4 — Getting Tables Out of R {#sec-export}

A table that only exists in an HTML preview is not a deliverable.

## Exercise 4.1 — Three backends

**Task.** Render `tbl_demog_final` through each engine below. Note one thing each does that the others do not.

```{r}
#| label: ex-4-1
#| eval: false

tbl_demog_final |> as_gt()
tbl_demog_final |> as_flex_table()
tbl_demog_final |> as_hux_table()
```

## Exercise 4.2 — RTF

**Task.** Export the demographics table to RTF: landscape, a title block carrying the protocol number and page numbering, a footnote block, and font size 9.

```{r}
#| label: ex-4-2
#| eval: false

library(huxtable)
library(pharmaRTF)

ht <- tbl_demog_final |>
  as_hux_table()

doc <- rtf_doc(ht) |>
  add_titles(
    hf_line(____, ____, align = "split", bold = TRUE),
    hf_line("____")
  ) |>
  add_footnotes(
    hf_line("____")
  ) |>
  set_font_size(____) |>
  set_orientation("____")

write_rtf(doc, file = "____.rtf")
```

::: {.callout-note collapse="true"}
## Hint 1

An `hf_line()` with `align = "split"` takes two strings and pushes them to opposite margins — protocol on the left, page number on the right.

Page numbering uses RTF field codes. Look at `?hf_line` for the tokens.
:::

::: callout-tip
## Check your work

Open the RTF in Word. Landscape, all columns on one page, page number correct on page 2.
:::

## Exercise 4.3 — Discussion

What would you add to make these programs acceptable in a validated environment? Consider dependency pinning, version stability across upgrades, independent QC, deterministic sorting, and traceability back to the source ADaM.

------------------------------------------------------------------------

# Reference Card {#sec-reference}

## Statistic placeholders

| Placeholder      | Meaning                  |
|------------------|--------------------------|
| `{n}`            | Count                    |
| `{N}`            | Denominator              |
| `{p}`            | Percent                  |
| `{N_nonmiss}`    | Non-missing count        |
| `{N_miss}`       | Missing count            |
| `{mean}`, `{sd}` | Mean, standard deviation |
| `{median}`       | Median                   |
| `{p25}`, `{p75}` | Quartiles                |
| `{min}`, `{max}` | Range                    |

## Modification functions

| Goal                           | Function                   |
|--------------------------------|----------------------------|
| Rename a column header         | `modify_header()`          |
| Add a spanning header          | `modify_spanning_header()` |
| Add a title                    | `modify_caption()`         |
| Add a footnote below the table | `modify_source_note()`     |
| Bold variable labels           | `bold_labels()`            |
| Add a total column             | `add_overall()`            |
| Show statistic names in rows   | `add_stat_label()`         |
| Stack tables vertically        | `tbl_stack()`              |
| Place tables side by side      | `tbl_merge()`              |
| Split and recombine            | `tbl_strata()`             |

## Selectors

`all_continuous()`, `all_categorical()`, `all_dichotomous()`, `all_stat_cols()`, `all_of()`, `starts_with()`, `everything()`

## Where to look things up

- Function reference and gallery: <https://www.danieldsjoberg.com/gtsummary/>
- `?tbl_summary` — every argument, examples at the foot of the page
- `show_header_names(tbl)` — the fastest way to find what `modify_header()` should target
